---
title: "Languages Of The World"
author: "Zakaria Laraki, Luc Dinh"
format: html
---

## Introduction: The Global Pulse of Language: Mapping Diversity and the Risk of Extinction
- Language is more than a tool for communication, it is a repositoty of history, culture, and unique cognitive perspectives. While globalization has pushed a handful of languages to the forefront of international trade and digital communication, thousands of others are now facing the threat of extinction. 

- In this project, we analyze the "Languages of the World" dataset to explore the spatial familial distribution of languages across the world. Our research focuses on identifying linguistic diversity hotspots and understanding the structural differences between widespead language families and those on the verge of vanishing. We aim to move beyond simple counts to inverstigae whether geographic isolation or family structure correlates with language endangerment.

- Ultimately, we want to answer the following questions: 
  - How is linguistic diversity geographically distributed across macro-areas? 
  - Does the size or isolation of a language family serve as a predictor for endangerment? 
  - What is the relationship between a language's geographic spread and its current status? 

## Data
In this project, we adopt different libraries to process and visualize the dataset


```{r}
# Load tidyverse package to help tidy and clean data sets
library("tidyverse")

# Used for visualizaing and plotting the dataset
library(plotly)
library(sf)

```
The primary source of our data is "The Languages of the World" dataset, curated from Glottolog 5.2.1, an open-access database in linguistics, maintained by the Max Planck Institute for Evolutionary Anthropology. Glottolog is the most comprehensive language database in linguistics, and contains information (names, genealogy, geographical information, endangerment status, etc.) of over 8,000 languages of the world. The database provide crucial linguistic information about the current status and diversity of the language in the world.

The data initially consists of 3 datasets (endangered_status, families, and languages). The endangered_status contain 8567 observations and 3 variables, within the endangered_status, the status variables contains 6 value representing the current status of the languages based of their IDs. we also have 4832 observations with 2 variables for Families, 7418 observations of 11 variables for languages. These individuals files need to be combined into one structured data frame for better analysis, plotting and require in-depth cleaning to ensure that no missing values will affects the insights derived from it

In this project, I read directly from the url links provided by the site github. I read and store the source data into 3 separate csv files
The link to the site can be found here:  https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-12-23/readme.md

```{r}
# Read and store the data from the source
endangered_status <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-23/endangered_status.csv')

families <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-23/families.csv')

languages <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-12-23/languages.csv')


```

```{r}
summary(families)
```

We started the data cleaning process by first joining the 3 csv files in to one single, structured variables "languages_clean" using comming variable of "id" and "family_id"
Next, we filtered the dataset for non null value in the macroarea and status_label variable then convert several different variables into factor for better grouping and plotting
Finally, we dropped any remaining rows containing NA in the dataset
```{r}
languages_clean <- languages |>
  # Join  datasets
  left_join(endangered_status, by = "id") |>
  left_join(families, by = c("family_id" = "id")) |>
  
  # Filter for specific non-NA values 
  filter(!is.na(macroarea), !is.na(status_label)) |>
  
  # Convert columns to factors
  mutate(
    macroarea = as.factor(macroarea),
    family    = as.factor(family),
    countries = as.factor(countries),
    
    # Group together
    status_group = case_when(
      status_label == "not endangered" ~ "Safe",
      status_label %in% c("shifting", "threatened") ~ "At risk",
      status_label %in% c("moribund", "nearly extinct") ~ "Severely endangered",
      status_label == "extinct" ~ "Extinct"
    )
  ) 

summary(languages_clean)
```


```{r}
languages_clean |>
  count(status_group, sort = TRUE) |>
  plot_ly(
    x = ~status_group, 
    y = ~n, 
    type = "bar", 
    marker = list(color = "darkgreen")
  ) |>
  layout(
    title = "Distribution of Language Endangerment Status",
    xaxis = list(title = "Number of Languages"),
    yaxis = list(title = "Status")
  )
```


```{r}
# Check at risk languages percentage
languages_clean |>
  mutate(at_risk = status_group != "Safe") |>
  count(at_risk) |>
  mutate(percent = n / sum(n) * 100)
```
### Core finding 
- Approximately $66%$ of languages are at risk(threatened, shifting, moribund, nearly extinct, or extinct) while only about $34%$ are not endagered. 
- This highlights that the majority of the world's languages are experiencing some level of decline. 

```{r}

plot_ly( data = languages_clean,
         x = ~macroarea,
         type = "histogram",
         marker = list(color = 'lightgreen',
                       line = list(color = 'black', width = 1)))|> 
  layout (
           title = "Languages by Macroarea",
           xaxis = list(title = "Macroarea"),
           yaxis = list(title = "Number of Languages"),
           margin = list(b = 100))
         
```




```{r}
plot_ly( data = languages_clean,
         x = ~is_isolate,
         type = "histogram",
         marker = list(color = 'steelblue',
                       line = list(color = 'black', width = 1)))|> 
  layout (
           title = "language isolated",
           xaxis = list(title = "Isolated"),
           yaxis = list(title = "Number of Languages"),
           margin = list(b = 100))
```
```{r}
top_5_language <- languages_clean |>
  separate_rows(countries, sep = ", ") |> 
  count(countries, name = "language_count") |>
  slice_max(order_by = language_count, n = 5) |>
  arrange(desc(language_count))
         
plot_ly(
  data = top_5_language,
  x = ~reorder(countries, -language_count),
  y = ~language_count,
  type = "bar",
  text = ~language_count,
  textposition = 'auto',
  marker = list(color = 'rgba(55, 128, 191, 0.7)',
                line = list(color = 'rgba(55, 128, 191, 1.0)', width = 2))
) |>
  layout(
    title = "Top 5 Countries with Highest Language Diversity",
    xaxis = list(title = "Country Code"),
    yaxis = list(title = "Number of Languages"),
    margin = list(t = 50)
  )
```

```{r}
world_sf <- read_sf("https://raw.githubusercontent.com/nvkelso/natural-earth-vector/master/geojson/ne_110m_admin_0_countries.geojson")

country_diversity <- languages_clean |>
  separate_rows(countries, sep = "[,;]\\s*") |>
  mutate(countries = toupper(trimws(as.character(countries)))) |> 
  count(countries, name = "language_count")

top_5_list <- country_diversity |> 
  slice_max(order_by = language_count, n = 5) |> 
  pull(countries)

world_map_final <- world_sf |>
  rename_with(tolower) |> 
  mutate(iso_a2 = toupper(iso_a2)) |> 
  left_join(country_diversity, by = c("iso_a2" = "countries")) |>
  mutate(status = ifelse(iso_a2 %in% top_5_list, "Top 5 Diverse", "Other"))

ggplot(data = world_map_final) +
  geom_sf(aes(fill = language_count), color = "white", size = 0.1) +
  geom_sf_text(aes(label = name), 
               size = 1.5, 
               color = "gray20", 
               check_overlap = TRUE) + 
  scale_fill_viridis_c(option = "mako", name = "Count") +
  theme_void()
```


## Analysis
### How is linguistic diversity geographically distributed? 
- Lets examine how the patterns vary geographically

```{r}
languages_clean |>
  count(macroarea, status_group) |>
  group_by(macroarea) |>
  mutate(prop = n / sum(n)) |>
  plot_ly(
    x = ~macroarea,
    y = ~prop,
    color = ~status_group,
    type = "bar", 
    colors = "Dark2" 
  ) |>
  layout(
    title = "Endangerment Distribution Across Macroareas",
    xaxis = list(title = "Macroarea"),
    yaxis = list(title = "Proportion"),
    barmode = "stack"
  )
```
- The distribution of language endangerment varies significantly across macroareas. Regions such as Africa: Eurasia and Africa: Eurasia: North America show a high proportion of safe languages, indicating relatively stable linguistic environments. In contrast, regions like Australia and parts of Eurasia display a much higher concentration of extinct and severly endangered languages, suggesting severe language loss. Additionally, areas such as Papunesia and South America exhibit a more balanced distribution across categories, with many languages at risk, indicating ongoing but not yet complete language decline.

- These patterns suggest that geographic location and regional context play an important role in language survival, supporting the idea that linguistic diversity and endangerment are unevenly distributed across the world.

### Are language isolates more endangered than languages that belond to families? 
```{r}
languages_clean |>
  mutate(language_type = if_else(is_isolate, "Isolate", "Family")) |>
  count(language_type, status_group) |>
  group_by(language_type) |>
  mutate(prop = n / sum(n)) |>
  plot_ly(
    x = ~status_group,
    y = ~prop,
    color = ~language_type,
    type = "bar"
  ) |>
  layout(
    title = "Endangerment Proportions: Isolates vs Families",
    xaxis = list(title = "Status"),
    yaxis = list(title = "Proportion of Languages"),
    barmode = "group"
  )
```
- The comparison between language isolates and languages within families reveals a clear difference in vulnerability. Using proportions instea than raw counts allows for a fair comparison, since isolates represent a much smaller group. Isolates show a higher proportion of languages classified as “at risk” or “extinct,” and a significantly lower proportion of “safe” languages. In contrast, languages that belong to families are more likely to remain stable, with a larger share classified as “not endangered.”

- This suggests that belonging to a language family may provide some resilience against extinction, possibly due to shared linguistic structure, cultural reinforcement, or broader speaker communities. Conversely, isolates, which lack related languages, appear to be more vulnerable to decline and disappearance.

- Also, This represents the strongest evidence in our analysis that structural linguistic factors, such as family membership, may influence language survival.

### now lets examine this using a map
```{r}
plot_ly(
  data = languages_clean |> filter(status_group == "Severely endangered"),
  lat = ~latitude,
  lon = ~longitude,
  type = "scattergeo",
  mode = "markers",
  text = ~name,
  marker = list(size = 4, color = "red", opacity = 0.7)
) |>
  layout(
    title = "Global Distribution of Severely Endangered Languages",
    geo = list(
      showland = TRUE,
      landcolor = "rgb(240,240,240)",
      showcountries = TRUE
    )
  )
```
- The spatial distribution of severely endangered languages reveals clear geographic clustering rather than a uniform global spread. High concentrations appear in regions such as Australia, Papua New Guinea, parts of Southeast Asia, and the western coasts of North and South America. These areas are known for high linguistic diversity, but also for smaller, more isolated language communities.

- This pattern suggests that regions with historically high diversity are also more vulnerable to language loss, likely due to factors such as globalization, cultural assimilation, and limited speaker populations. In contrast, regions dominated by widely spoken languages show fewer severely endangered languages, highlighting the role of geographic and social concentration in language survival.

- The spatial clustering observed here reinforces earlier findings that vulnerability is not random, but concentrated in specific high-diversity regions

## Results


## Conclusion
